---
title: "Contributing to nflfastR"
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  comment = "#>"
)

patch_columns <- yaml::read_yaml("data-raw/patch_valid_column_names.yaml")$column
table_data <- tibble::tibble(
  a = patch_columns[1:60],
  b = patch_columns[61:120],
  c = patch_columns[121:180],
  d = c(patch_columns[181:234], rep(NA_character_, 6))
)

```

Thanks for your interest in contributing! There are many ways to make a valuable contribution to an open-source project, such as issues or pull requests.

This guide focuses specifically on how users (or AI agents) can make specific data corrections if the raw data contains minor errors.

Basically, there are two possible reasons why data in nflfastR might be incorrect:

1. The parser has a bug and does not handle a specific situation correctly
1. The raw data is incorrect

In the first case, please submit issues or pull requests that explain the problem and, if necessary, propose changes to the parser.

For the second case, nflfastR provides a mechanism for manually overwriting the data before the parser applies its usual modifications to the raw data.

## Contribute manual data corrections

Manual corrections should be the last resort when we have no other way to fix an error. Alternatively, we can also use them if a fix in the parser would be very time-consuming or potentially prone to errors. A typical example is rare errors in the raw data, particularly in the raw data from older seasons. So if there are only a handful of incorrect values, this might be the right approach.

Here's how to submit manual corrections:

1. fork the repo
1. create new branch (we have no naming conventions)
1. add patch data to `"inst/patch_pbp.json"` and save the file
1. run  `devtools::load_all()`
1. run `.patch_clean_data()` interactively to ensure correct sorting and to remove duplicates
1. run `devtools::test(filter = "patch_pbp")`. This will run tests aimed at the file format of the patch file in `"inst/patch_pbp.json"`.
1. run `build_nflfastR_pbp()` with the `game_id`s  affected by your correction to confirm that your patch works as intended
1. commit and push
1. create the pull request

A patch in the `"inst/patch_pbp.json"` file looks like this

```json
[
  {
    "game_id": "2015_01_IND_BUF",
    "play_id": 1207,
    "column": "quarter",
    "value": [2]
  },
  {
    "game_id": "2014_07_ATL_BAL",
    "play_id": 99999,
    "column": "start_time",
    "value": ["10/19/14, 13:02:00"]
  }
]
```

Some notes:

- The first entry overwrites the `qtr` variable (special name, see table 1) in the play-by-play data of `game_id 2015_01_IND_BUF` and `play_id 1207` with the integer `2`.
- The second entry overwrites the `start_time` variable in the play-by-play data of the complete `game_id 2014_07_ATL_BAL` with the string `"10/19/14, 13:02:00"`.
- Please note the `[]` brackets in the value field. They allow us to save `value` in different classes, i.e. `integer`, `numeric`, `character`. We apply validation to ensure that this value is a list of length one. 
- Integers and numerics don't need quotes. We want to preserve their classes.
- The field `play_id` is also unquoted and should be an integer. 
- The special `play_id 99999` is a placeholder and tells nflfastR to apply the patch to the complete column, i.e. all `play_id`s of the specified `game_id`. 
- Only `value` is allowed to be `NA`, which means you would like to overwrite specifically with `"NA"`.
- Valid column names to be patched are listed in table 2 below.
- There are five more or less special column names that are renamed in the parsing process (that's a legacy thing we inherited from nflscrapR for backwards compatibility). This means that five column names are different in the patch process than in the final pbp. The following Table 1 shows the column names used in the final pbp and how they must be named in the patch file.

**Table 1: Column names in final `pbp` matched to their name in the `column` field of `"inst/patch_pbp.json"`**

```{r}
knitr::kable(
  tibble::tribble(
    ~pbp            , ~patch_file        ,
    "qtr"       , "quarter"          ,
    "yrdln"         , "yardline"         ,
    "ydstogo"       , "yards_to_go"      ,
    "side_of_field" , "yardline_side"    ,
    "desc"          , "play_description"
  ),
  col.names = c("pbp name", "patch file name")
)
```


Package tests are designed to enforce the above described format criteria.

```{r eval=FALSE}
DT::datatable(
  table_data,
  options = list(scrollX = TRUE, pageLength = 25, dom = "ftip", ordering = FALSE),
  filter = "none",
  rownames = FALSE,
  colnames = NULL,
  style = "auto"
)
```

**Table 2: Column names allowed in the `column` field of `"inst/patch_pbp.json"`**

```{r}
options(knitr.kable.NA = '---')
knitr::kable(table_data, col.names = rep("", 4))
```

