---
title: "Get started with nflfastR"
author: "Ben Baldwin & Sebastian Carl"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
future::plan("multisession")
options(dplyr.summarise.inform = FALSE)
options(nflreadr.verbose = FALSE)
```

If you are new to R or are having trouble understanding the code in the below sections we highly recommend the **nflfastR beginner's guide** in `vignette("beginners_guide")`.

# The Main Functions

nflfastR comes with a set of functions to access NFL play-by-play data and team rosters. This section provides a brief introduction to the essential functions.

nflfastR processes and cleans up play-by-play data and adds variables through [it's models](https://www.opensourcefootball.com/posts/2020-09-28-nflfastr-ep-wp-and-cp-models/). Since some of these tasks are performed by separate functions, the easiest way to compute the complete nflfastR dataset is `build_nflfastR_pbp()`. The main input for that function is a set of game ids which can be accessed with `fast_scraper_schedules()`. The following code demonstrates how to build the nflfastR dataset for the Super Bowls of the 2017 - 2019 seasons.

```{r}
library(nflfastR)
library(dplyr, warn.conflicts = FALSE)
ids <- nflfastR::fast_scraper_schedules(2017:2019) |>
  dplyr::filter(game_type == "SB") |>
  dplyr::pull(game_id)
pbp <- nflfastR::build_nflfastR_pbp(ids)
```

In most cases, however, it is not necessary to use this function for individual games, because nflfastR provides both a [data release](https://github.com/nflverse/nflverse-data/releases/tag/pbp) and two main play-by-play functions: `load_pbp()` and `update_db()`. We cover `load_pbp()` below, and please see [Example 8: Using the built-in database function] for how to work with the database function `update_db()`.

The easiest way to access the data from the release is the function `load_pbp()`. It can load multiple seasons directly into memory and supports multiple data formats. Loading all play-by-play data of the 2022-2024 seasons is as easy as 

```{r}
pbp <- nflfastR::load_pbp(2022:2024)
```

Joining roster data to the play-by-play data set is possible as well. The data can be accessed with the function `fast_scraper_roster()` and its application is demonstrated in [Example 10: Working with roster and position data].

# Application Examples

All examples listed below assume that the following two libraries are installed and loaded.

``` {r load, warning = FALSE, message = FALSE}
library(nflfastR)
library(tidyverse)
```

## Example 1: Completion Percentage Over Expected (CPOE)

Let's look at CPOE leaders from the 2009 regular season.

As discussed above, `nflfastR` has a data release for all available seasons, so there's no need to actually scrape them. Let's use that here with the convenience function `load_pbp()` which fetches data from the release (for non-R users, .csv and .parquet are also available in the [data release](https://github.com/nflverse/nflverse-data/releases/tag/pbp)).

``` {r ex3-cpoe, warning = FALSE, message = FALSE}
games_2009 <- nflfastR::load_pbp(2009) |> dplyr::filter(season_type == "REG")
games_2009 |>
  dplyr::filter(!is.na(cpoe)) |>
  dplyr::summarize(
    passer = nflreadr::stat_mode(passer_player_name),
    cpoe = mean(cpoe), 
    Atts = n(), 
    .by = passer_player_id
  ) |>
  dplyr::filter(Atts > 200) |>
  dplyr::slice_max(cpoe, n = 5) |>
  knitr::kable(digits = 1)
```

## Example 2: Using Drive Information

When working with `nflfastR`, drive results are automatically included. We use `fixed_drive` and `fixed_drive_result` since the NFL-provided information is a bit wonky. Let's look at how much more likely teams were to score starting from 1st & 10 at their own 20 yard line in 2015 (the last year before touchbacks on kickoffs changed to the 25) than in 2000.

``` {r ex4, warning = FALSE, message = FALSE}
pbp <- nflfastR::load_pbp(c(2003, 2015))

out <- pbp |>
  dplyr::filter(season_type == "REG" & down == 1 & ydstogo == 10 & yardline_100 == 80) |>
  dplyr::mutate(drive_score = dplyr::if_else(fixed_drive_result %in% c("Touchdown", "Field goal"), 1, 0)) |>
  dplyr::summarize(drive_score = mean(drive_score), .by = season)

out |> 
  knitr::kable(digits = 3)
```

So `r scales::percent(out$drive_score[1], accuracy = 0.1)` of 1st & 10 plays from teams' own 20 would see the drive end up in a score in 2003, compared to `r scales::percent(out$drive_score[2], accuracy = 0.1)` in 2015. This has implications for Expected Points models (see [this article](https://www.opensourcefootball.com/posts/2020-09-28-nflfastr-ep-wp-and-cp-models/)).

## Example 3: Plot offensive and defensive EPA per play for a given season

Let's build the **[NFL team tiers](https://rbsdm.com/stats/stats/)** using offensive and defensive expected points added per play for the 2005 regular season. Creating data viz including NFL team logos (or wordmarks, or headshots), we recommend the nflverse R package [nflplotR](https://nflplotr.nflverse.com).

When using `load_pbp()`, the helper function `clean_pbp()` has already been run, which creates "rush" and "pass" columns that (a) properly count sacks and scrambles as pass plays and (b) properly include plays with penalties. Using this, we can keep only rush or pass plays.

```{r ex5, warning = FALSE, message = FALSE, results = 'hide', fig.keep = 'all', dpi = 600}
library(nflplotR)
pbp <- nflfastR::load_pbp(2005) |>
  dplyr::filter(season_type == "REG") |>
  dplyr::filter(!is.na(posteam) & (rush == 1 | pass == 1))
offense <- pbp |>
  dplyr::group_by(team = posteam) |>
  dplyr::summarise(off_epa = mean(epa, na.rm = TRUE))
defense <- pbp |>
  dplyr::group_by(team = defteam) |>
  dplyr::summarise(def_epa = mean(epa, na.rm = TRUE))
offense |>
  dplyr::inner_join(defense, by = "team") |>
  ggplot2::ggplot(aes(x = off_epa, y = def_epa)) +
  ggplot2::geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3), alpha = .2) +
  nflplotR::geom_mean_lines(aes(y0 = off_epa, x0 = def_epa)) +
  nflplotR::geom_nfl_logos(aes(team_abbr = team), width = 0.07, alpha = 0.7) +
  ggplot2::labs(
    x = "Offense EPA/play",
    y = "Defense EPA/play",
    caption = "Data: @nflfastR",
    title = "2005 NFL Offensive and Defensive EPA per Play"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 12, hjust = 0.5, face = "bold")
  ) +
  ggplot2::scale_y_reverse()
```

## Example 4: Expected Points calculator

We have provided a calculator for working with the Expected Points model. Here is an example of how to use it, looking for how the Expected Points on a drive beginning following a touchback has changed over time.

While I have put in `'SEA'` for `home_team` and `posteam`, this only matters for figuring out whether the team with the ball is the home team (there's no actual effect for given team; it would be the same no matter what team is supplied).

``` {r ex6a}
data <- tibble::tibble(
  "season" = 1999:2019,
  "home_team" = "SEA",
  "posteam" = "SEA",
  "roof" = "outdoors",
  "half_seconds_remaining" = 1800,
  "yardline_100" = c(rep(80, 17), rep(75, 4)),
  "down" = 1,
  "ydstogo" = 10,
  "posteam_timeouts_remaining" = 3,
  "defteam_timeouts_remaining" = 3
)
nflfastR::calculate_expected_points(data) |>
  dplyr::select(season, yardline_100, td_prob, ep) |>
  knitr::kable(digits = 2)
```

Not surprisingly, offenses have become much more successful over time, with the kickoff touchback moving from the 20 to the 25 in 2016 providing an additional boost. Note that the `td_prob` in this example is the probability that the next score within the same half will be a touchdown scored by team with the ball, **not** the probability that the current drive will end in a touchdown (this is why the numbers are different from Example 4 above).

We could compare the most recent four years to the expectation for playing in a dome by inputting all the same things and changing the `roof` input:

``` {r ex6b}
data <- tibble::tibble(
  "season" = 2016:2019,
  "week" = 5,
  "home_team" = "SEA",
  "posteam" = "SEA",
  "roof" = "dome",
  "half_seconds_remaining" = 1800,
  "yardline_100" = c(rep(75, 4)),
  "down" = 1,
  "ydstogo" = 10,
  "posteam_timeouts_remaining" = 3,
  "defteam_timeouts_remaining" = 3
)
nflfastR::calculate_expected_points(data) |>
  dplyr::select(season, yardline_100, td_prob, ep) |>
  knitr::kable(digits = 2)
```

So for 2018 and 2019, 1st & 10 from a home team's own 25 yard line had higher EP in domes than at home, which is to be expected.

## Example 5: Win probability calculator

We have also provided a calculator for working with the win probability models. Here is an example of how to use it, looking for how the win probability to begin the game depends on the pre-game spread.

While I have put in `'SEA'` for `home_team` and `posteam`, this only matters for figuring out whether the team with the ball is the home team (there's no actual effect for given team; it would be the same no matter what team is supplied).

``` {r ex7}
data <- tibble::tibble(
  "receive_2h_ko" = 0,
  "home_team" = "SEA",
  "posteam" = "SEA",
  "score_differential" = 0,
  "half_seconds_remaining" = 1800,
  "game_seconds_remaining" = 3600,
  "spread_line" = c(1, 3, 4, 7, 14),
  "down" = 1,
  "ydstogo" = 10,
  "yardline_100" = 75,
  "posteam_timeouts_remaining" = 3,
  "defteam_timeouts_remaining" = 3
)
nflfastR::calculate_win_probability(data) |>
  dplyr::select(spread_line, wp, vegas_wp) |>
  knitr::kable(digits = 2)
```

Not surprisingly, `vegas_wp` increases with the amount a team was coming into the game favored by.

## Example 6: Using the built-in database function

If you're comfortable using `dplyr` functions to manipulate and tidy data, you're ready to use a database. Why should you use a database?

* The provided function in `nflfastR` makes it extremely easy to build a database and keep it updated
* Play-by-play data over 20+ seasons takes up a lot of memory: working with a database allows you to only bring into memory what you actually need
* R makes it *extremely* easy to work with databases.

### Start: install and load packages

To start, we need to install the two packages required for this that aren't installed automatically when `nflfastR` installs: `DBI` and `RSQLite` (advanced users can use other types of databases, but this example will use SQLite). The `if` statements make sure the packages won't be updated if they are already installed:

``` {r eval = FALSE}
if (!require("DBI")) install.packages("DBI")
if (!require("RSQLite")) install.packages("RSQLite")
```

As with always, you only need to install these once. They don't need to be loaded to build the database because `nflfastR` knows how to use them, but we do need them later on when working with the database.

``` {r}
library(DBI)
library(RSQLite)
```

### Build database

There's exactly one function in `nflfastR` that works with databases: `update_db()`. Some notes:

* If you use `update_db()` with no arguments, it will build a SQLite database called `pbp_db` in your current working directory, with play-by-play data in a table called `nflfastR_pbp`.
* You can specify a different directory with `dbdir`.
* You can specify a different filename with `dbname`.
* You can specify a different table name with `tblname`.
* If you want to rebuild the database from scratch for whatever reason, supply `force_rebuild = TRUE`. This is primarily intended for the case when we update the play-by-play data in the data repo due to fixing a bug and you want to force the database to be wiped and updated.
* If you want to rebuild specified seasons, this can also be supplied to `force_rebuild` (e.g. `force_rebuild = c(2019, 2020)`).
* The parameter `db_connection` is intended for advanced users who want to use other [DBI drivers](https://dbi.r-dbi.org/reference/dbidriver-class), such as MariaDB, Postgres or odbc. Please note that `dbdir` and `dbname` are dropped when a `db_connection` is provided but the argument `tblname` will still be used to write the data table into the database.

Let's say I just want to dump a database into the current working directory. Here we go!

``` {r create-db}
nflfastR::update_db()
```

This created a database in the current directory called `pbp_db`.

Wait, that's it? That's it! What if it's partway through the season and you want to make sure all the new games are added to the database? What do you run? `update_db()`! (just make sure you're in the directory the database is saved in or you supply the right file path)

``` {r update-db}
nflfastR::update_db()
```

If it's partway through a season and you want to re-build a season to allow for data corrections from the NFL to propagate into your database, you can specify one season to be rebuilt:

``` {r update-db-yr}
nflfastR::update_db(force_rebuild = 2020)
```

### Connect to database

Now we can make a connection to the database. This is the only part that will look a little bit foreign, but all you need to know is where your database is located. If it's in your current working directory, this will work:

``` {r}
connection <- DBI::dbConnect(RSQLite::SQLite(), "./pbp_db")
connection
```

It looks like nothing happened, but we now have a connection to the database. Now we're ready to do stuff. If you aren't familiar with databases, they're organized around tables. Here's how to see which tables are present in our database: 

``` {r}
DBI::dbListTables(connection)
```

Since we went with the defaults, there's a table called `nflfastR_pbp`. Another useful function is to see the fields (i.e., columns) in a table:

``` {r}
DBI::dbListFields(connection, "nflfastR_pbp") |>
  utils::head(10)
```

This is the same list as the list of columns in `nflfastR` play-by-play. Notice we had to supply the name of the table above (`"nflfastR_pbp"`). 

With all that out of the way, there's only a couple more things to learn. The main driver here is `tbl`, which helps get output with a specific table in a database:

``` {r}
pbp_db <- dplyr::tbl(connection, "nflfastR_pbp")
```

And now, everything will magically just "work": you can forget you're even working with a database!

``` {r}
pbp_db |>
  dplyr::group_by(season) |>
  dplyr::summarize(n = dplyr::n())
pbp_db |>
  dplyr::filter(rush == 1 | pass == 1, down <= 2, !is.na(epa), !is.na(posteam)) |>
  dplyr::group_by(pass) |>
  dplyr::summarize(mean_epa = mean(epa, na.rm = TRUE))
```

So far, everything has stayed in the database. If you want to bring a query into memory, just use `collect()` at the end:

``` {r}
russ <- pbp_db |>
  dplyr::filter(name == "R.Wilson" & posteam == "SEA") |>
  dplyr::select(desc, epa) |>
  dplyr::collect()
russ
```

So we've searched through about 1 million rows of data across 300+ columns and only brought about `r round(nrow(russ), -1)` rows and two columns into memory. Pretty neat! This is how I supply the data to the shiny apps on rbsdm.com without running out of memory on the server. Now there's only one more thing to remember. When you're finished doing what you need with the database:

``` {r}
DBI::dbDisconnect(connection)
```

For more details on using a database with `nflfastR`, see [Thomas Mock's life-changing post here](https://themockup.blog/posts/2019-04-28-nflfastr-dbplyr-rsqlite/). More detailed information on dbplyr (the dplyr database back-end) are given in the second edition of [Hadley Wickham's R for Data Science (2e)](https://r4ds.hadley.nz/import-databases.html).

## Example 7: working with the expected yards after catch model

The variables in `xyac` are as follows:

* `xyac_epa`: The expected value of EPA gained after the catch, **starting from where the catch was made**.
* `xyac_success`: The probability the play earns positive EPA (relative to where play started) based on where ball was caught.
* `xyac_fd`: Probability play earns a first down based on where the ball was caught.
* `xyac_mean_yardage` and `xyac_median_yardage`: Average and median expected yards after the catch based on where the ball was caught.

Some other notes:

* `epa` = `air_epa` + `yac_epa`, where `air_epa` is the EPA associated with a catch at the target location. If a receiver loses a fumble, it is removed from his `yac_epa`
* Expected value of EPA at catch point = `air_epa` + `xyac_epa`
* So if we want to get YAC EPA over expected, we need to compare `yac_epa` to `xyac_epa`, as in the example below
* To get first downs over expected, we could compare `first_down` to `xyac_fd`
* These fields are populated for all pass attempts, whether caught or not, but restrict to completed passes when measuring, for example, YAC EPA over expected
* The expected YAC EPA model doesn't take receiver fumbles into account, so actual minus expected YAC is slightly negative due to fumbles happening

Let's create measures for EPA and first downs over expected in 2015:

``` {r ex9-xyac, warning = FALSE, message = FALSE}
nflfastR::load_pbp(2015) |>
  dplyr::group_by(receiver, receiver_id, posteam) |>
  dplyr::mutate(tgt = sum(complete_pass + incomplete_pass)) |>
  dplyr::filter(tgt >= 50) |>
  dplyr::filter(complete_pass == 1, air_yards < yardline_100, !is.na(xyac_epa)) |>
  dplyr::summarize(
    epa_oe = mean(yac_epa - xyac_epa),
    actual_fd = mean(first_down),
    expected_fd = mean(xyac_fd),
    fd_oe = mean(first_down - xyac_fd),
    rec = dplyr::n()
  ) |>
  dplyr::ungroup() |>
  dplyr::select(receiver, posteam, actual_fd, expected_fd, fd_oe, epa_oe, rec) |>
  dplyr::slice_max(epa_oe, n = 10) |>
  knitr::kable(digits = 3)
```

The presence of so many running backs on this list suggests that even though it takes into account target depth and pass direction, the model doesn't do a great job capturing space. Alternatively, running backs might be better at generating yards after the catch since running with the football is their primary role.

## Example 8: Working with roster and position data

At long last, there's a way to merge the new play-by-play data with roster information. Use the function to get the rosters:

``` {r roster}
roster <- nflfastR::fast_scraper_roster(2019)
```

Now let's load play-by-play data from 2019:
``` {r roster_pbp_load}
games_2019 <- nflfastR::load_pbp(2019)
```

Here is what the player IDs look like because `nflfastR` now automatically decodes IDs to look like the old format with GSIS IDs:

``` {r roster_pbp}
games_2019 |>
  dplyr::filter(rush == 1 | pass == 1, posteam == "SEA") |>
  dplyr::select(name, id)
```

Now we're ready to join to the roster data using these IDs:
``` {r decode_join}
joined <- games_2019 |>
  dplyr::filter(!is.na(receiver_id)) |>
  dplyr::select(posteam, season, desc, receiver, receiver_id, epa) |>
  dplyr::left_join(roster, by = c("receiver_id" = "gsis_id"))
```

``` {r decode_table}
# the real work is done, this just makes a table and has it look nice
joined |>
  dplyr::filter(position %in% c("WR", "TE", "RB")) |>
  dplyr::group_by(receiver_id, receiver, position) |>
  dplyr::summarize(tot_epa = sum(epa), n = n()) |>
  dplyr::arrange(-tot_epa) |>
  dplyr::ungroup() |>
  dplyr::group_by(position) |>
  dplyr::mutate(position_rank = 1:n()) |>
  dplyr::filter(position_rank <= 5) |>
  dplyr::rename(Pos_Rank = position_rank, Player = receiver, Pos = position, Tgt = n, EPA = tot_epa) |>
  dplyr::select(Player, Pos, Pos_Rank, Tgt, EPA) |>
  knitr::kable(digits = 0)
```

Not surprisingly, all 5 of the top 5 WRs in terms of EPA added come in ahead of the top RB. Note that the number of targets won't match official stats because we're including plays with penalties.

## Example 9: Replicating official stats

The columns like `name`, `passer`, `fantasy` etc are `nflfastR`-created columns that mimic "real" football: i.e., excluding plays with spikes, counting scrambles and sacks as pass plays, etc. But if you're trying to replicate official statistics -- perhaps for fantasy purposes -- use the `*_player_name` and `*_player_id` columns.

### Leaderboards

[Let's try to replicate this page of passing leaders](https://www.nfl.com/stats/player-stats/).

``` {r stats1}
nflfastR::load_pbp(2020) |>
  dplyr::filter(season_type == "REG", complete_pass == 1 | incomplete_pass == 1 | interception == 1, !is.na(down)) |>
  dplyr::group_by(passer_player_name, posteam) |>
  dplyr::summarize(
    yards = sum(passing_yards, na.rm = T),
    tds = sum(touchdown == 1 & td_team == posteam),
    ints = sum(interception),
    att = dplyr::n()
  ) |>
  dplyr::arrange(-yards) |>
  utils::head(10) |>
  knitr::kable(digits = 0)
```

These match the official stats on NFL.com (note the filter for `season_type == "REG"` since official stats only count regular season games). Note that we're using `passing_yards` here because `yards_gained` is not equal to passing yards on plays with laterals.

While the above code works in this case, there are several special cases where it is nearly impossible to get official player stats from nflfastR play-by-play data. The reason for this is that the idea of nflfastR play-by-play data is a “tidy” data structure. In other words, the aim is to have one row per play in the data. This can lead to problems if, for example, there are several changes of possession per play (i.e. several fumbles) or if the ball is lateraled in a play. These are just two examples of “abnormal” plays that are not fully captured in a tidy data structure.
We have solved this problem with the function `calculate_stats()`. This function uses playstats of the raw play-by-play data before it is parsed into a tidy structure by nflfastR. 

This function has the following features:

- It determines stats in offense, defense, and special teams,
- either on player level or on team level,
- and can summarize them on season level (separately for regular season and post season) or on week level.

For more information see the function documentation of `calculate_stats()`. Again, **don't try to get an exact match with official stats based on nflfastR play-by-play data**. It usually works, but fails because of details that are unsolvable.

Now let's replicate the above table using `calculate_stats()`:

``` {r stats2}
s <- nflfastR::calculate_stats(
  seasons = 2020, 
  summary_level = "season",
  stat_type = "player",
  season_type = "REG"
)
s |>
  dplyr::slice_max(passing_yards, n = 10) |> 
  dplyr::select(player_name, recent_team, completions, attempts, passing_yards, passing_tds, passing_interceptions, attempts) |>
  knitr::kable(digits = 0)
```

# Frequent issues

## The `drive` column looks wacky

Use `fixed_drive` and `fixed_drive_result` instead. See [Example 2: Using Drive Information].

## Why are there so many win probability columns?

`vegas_wp` and `vegas_home_wp` incorporate the pregame spread and are much better models.

## Need more help?

Please ask [in the Discord channel](https://discord.com/invite/5Er2FBnnQa).
